\sys: Automatic Disease Diagnosis Using Large Language Models and Answer Set Programming

### 2025-09-19 paper/draft.tex
Small tweaks
Signed-off-by: Evangelos Lamprou <vagos@lamprou.xyz>

In clinical decision support, the real goal is to get the best of both worlds: 
systems that reason accurately while still being easy for practitioners to trust and use. 
\sys takes a step in that direction by combining the strengths of LLMs and ASP. 
Instead of requiring experts to hand-craft large knowledge bases, 
\sys automatically converts medical literature into ASP rules. 
These rules are then enriched with patient symptoms and clinical indicators, 
and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.

Concretely, this work:
\begin{itemize}
    \item Introduces a framework that turns medical text into an executable ASP knowledge base.
    \item Uses prompts to generate modular rules (including alternatives and symptom links) that the solver applies for reasoning.
    \item Provides solver-based explanations showing how symptoms lead to rules and then to diagnoses.
    \item Demonstrates accurate results on real patient data while keeping the reasoning fully transparent.
\end{itemize}

On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
while giving transparent explanations for each diagnosis.

The paper begins with background on LLMs and ASP (\cref{sec:background}), 
then details the \sys methodology (\cref{sec:methodology}), 
evaluates the approach on patient data (\cref{sec:results}), 
and concludes with a discussion of limitations and future directions (\cref{sec:conclusion}).

@@

\begin{figure}[htb]
    \begin{center}
\begin{tikzpicture}[node distance=2.9cm]
        \node [simple-rect] (problem)  at (0,0) {Problem};
        \node [simple-rect, below of=problem] (program) {Program};

@@

\begin{figure}[!h]
    % https://docs.google.com/drawings/d/1cr2lTIjSUxfQH54CZui8vGqLS2keOfW41i4g4uGfwhw/edit?usp=sharing
    \includegraphics[width=\textwidth]{assets/mccoy.pdf}
    \caption{\textbf{\sys overview.} The diagnostic process begins by transforming medical literature 
    into an ASP program using a LLM. 
    An ASP solver then executes the program and produces a final diagnosis 
    based on each patient’s data.}
    \label{fig:architecture}
\end{figure}

@@

Symbolic Artifical Intelligence offers a significant advantage 
over subsymbolic methods in terms of explainability. 
In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
at the time of writing,\footnote{At the time of writing this paper.} 
provide intuitive traces of their reasoning processes. 
Prior work has explored methods for associating logical propositions 
with their justifications \cite{cabalar2014causal}. For example, 
the logic program shown below is accompanied by a causal graph 
in \cref{fig:causal-g}, which illustrates the connections 
between rules and their inferred outcomes. 
This structure supports the interpretability of conclusions derived by a symbolic logic solver.
% \begin{equation}
% \begin{aligned}[b]
%     l :& punish \longleftarrow drive, drunk \\
%     m :& punish \longleftarrow resist \\
%     e :& prison \longleftarrow punish \\
%     d :& drive \\
%     k :& drunk \\
%     r :& resist \\
% \end{aligned}
% \label{eq:example-cg}
% \end{equation}
% % https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
% \begin{figure}
%     \centering
%     \includegraphics{assets/causal_g.pdf}
%     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
%     \label{fig:causal-g}
% \end{figure}
\begin{figure}[!h]
    \centering
    \begin{minipage}{0.48\textwidth}
    \begin{equation}
    \begin{aligned}[b]
        l :& punish \longleftarrow drive, drunk \\
        m :& punish \longleftarrow resist \\
        e :& prison \longleftarrow punish \\
        d :& drive \\
        k :& drunk \\
        r :& resist \\
    \end{aligned}
    \label{eq:example-cg}
    \end{equation}
    \end{minipage}
    \hfill
    \begin{minipage}{0.48\textwidth}
    \centering
    \includegraphics[width=\textwidth]{assets/causal_g.pdf}
    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
    \label{fig:causal-g}
    \end{minipage}
    \end{figure}

@@

Future work will focus on grounding ASP rules in reliable sources 
through retrieval-augmented generation, fine-tuning LLMs for clinical reasoning, 
and scaling the framework to cover a wider range of diseases.
\newpage
\bibliographystyle{unsrt}
\bibliography{bib}

>>>>

In clinical decision support, the real goal is to get the best of both worlds: 
systems that reason accurately while still being easy for practitioners to trust and use. 
\sys takes a step in that direction by combining the strengths of LLMs and ASP. 
Instead of requiring experts to hand-craft large knowledge bases, 
\sys automatically converts medical literature into ASP rules. 
These rules are then enriched with patient symptoms and clinical indicators, 
and passed through an ASP solver to produce diagnostic suggestions together with logical justifications.
On public patient records, \sys reaches 95–100\% predictive accuracy on selected diseases, 
while giving transparent explanations for each diagnosis.

% Concretely, this work:
% \begin{itemize}
%     \item Introduces a framework that combines the flexibility of LLMs with the explanability of ASP for disease diagnosis.
%     \item Provides solver-based explanations showing how symptoms lead to rules and then to diagnoses.
%     \item Demonstrates accurate results on real patient data while keeping the reasoning fully transparent.
% \end{itemize}

The paper begins with background on LLMs and ASP (\cref{sec:background}), 
then details the \sys framework (\cref{sec:methodology}), 
evaluates the approach on patient data (\cref{sec:results}), 
and concludes with a discussion of limitations and future directions (\cref{sec:conclusion}).

@@

\begin{figure}[t]
    \begin{center}
\begin{tikzpicture}[node distance=2.9cm]
        \node [simple-rect] (problem)  at (0,0) {Problem};
        \node [simple-rect, below of=problem] (program) {Program};

@@

\begin{figure}[t]
    % https://docs.google.com/drawings/d/1cr2lTIjSUxfQH54CZui8vGqLS2keOfW41i4g4uGfwhw/edit?usp=sharing
    \includegraphics[width=\textwidth]{assets/mccoy.pdf}
    \caption{\textbf{\sys overview.} The diagnostic process begins by transforming medical literature 
    into an ASP program using a LLM. 
    An ASP solver then executes the program and produces a final diagnosis 
    based on each patient’s data.}
    \label{fig:architecture}
\end{figure}

@@

Symbolic Artifical Intelligence offers a significant advantage 
over subsymbolic methods in terms of explainability. 
In symbolic systems, the relationships between facts and conclusions are explicitly defined, 
enabling transparent reasoning. However, ASP solvers such as \textit{Clingo} do not, 
at the time of writing,\footnote{At the time of writing this paper.} 
provide intuitive traces of their reasoning processes. 
Prior work has explored methods for associating logical propositions 
with their justifications \cite{cabalar2014causal}. For example, 
the logic program shown below is accompanied by a causal graph 
in \cref{fig:causal-g}, which illustrates the connections 
between rules and their inferred outcomes. 
This structure supports the interpretability of conclusions derived by a symbolic logic solver.
% \begin{equation}
% \begin{aligned}[b]
%     l :& punish \longleftarrow drive, drunk \\
%     m :& punish \longleftarrow resist \\
%     e :& prison \longleftarrow punish \\
%     d :& drive \\
%     k :& drunk \\
%     r :& resist \\
% \end{aligned}
% \label{eq:example-cg}
% \end{equation}
% % https://docs.google.com/drawings/d/1Ing4GFH2kbniiHI5rbXzw-xcAUQAxWAJpgmg4tz1fzc/edit?usp=sharing
% \begin{figure}
%     \centering
%     \includegraphics{assets/causal_g.pdf}
%     \caption{Causal graph of logic program (\cref{eq:example-cg}).}
%     \label{fig:causal-g}
% \end{figure}
\begin{figure}[t]
    \centering
    \begin{minipage}{0.48\textwidth}
    \begin{equation}
    \begin{aligned}[b]
        l :& punish \longleftarrow drive, drunk \\
        m :& punish \longleftarrow resist \\
        e :& prison \longleftarrow punish \\
        d :& drive \\
        k :& drunk \\
        r :& resist \\
    \end{aligned}
    \label{eq:example-cg}
    \end{equation}
    \end{minipage}
    \hfill
    \begin{minipage}{0.45\textwidth}
    \centering
    \includegraphics[width=.5\textwidth]{assets/causal_g.pdf}
    \caption{Causal graph of logic program (\cref{eq:example-cg}).}
    \label{fig:causal-g}
    \end{minipage}
\end{figure}

@@

Future work will focus on grounding ASP rules in reliable sources 
through retrieval-augmented generation, fine-tuning LLMs for clinical reasoning, 
and scaling the framework to cover a wider range of diseases.
\bibliographystyle{unsrt}
\bibliography{bib}
